Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2014-10-28 AO_bondedSi.tex
overhaul the README.md file to include code and figure descriptions
complete overhaul of the introduction from DTJ email dated Sept 2, 2014

Direct optical bonding of silicon allows manufacturers to combine Si parts with no Fresnel loss at the interface.  The high refractive index of Si makes this advantage particularly useful.  Traditional figures of merit for the optical audits of bonds, however, are either insensitive to gaps that are small yet have a significant impact on throughput and phase coherence, or do not provide quantitative information about the extent of the gaps along the direction of propagation.

%Issues with the quality of the bonds become particularly acute when the optics are used in double-pass and/or when the incidence angle at the bonds is far from the normal.

The University of Texas diffractive optics group produces diffraction gratings on silicon using microlithographic techniques \cite{1998SPIE.3354..201J,2010SPIE.7739E.146W}.  These gratings are used either in transmission (i.e. as grisms \cite{2008SPIE.7014E..77D, 2010SPIE.7739E.123G}), or in reflection (i.e. as immersion gratings, \cite{2007ApOpt..46.3400M, 2010SPIE.7739E.146W,2012SPIE.8450E..2SG}.  These gratings operate throughout the infrared, from 1.2 to 38 $\mu$m, and find application in ground-based (IGRINS \cite{2010SPIE.7735E..54Y, 2012SPIE.8450E..2SG}), airborne (SOFIA FORCAST \cite{2008SPIE.7014E..77D}), and space (JWST NIRCAM \cite{2005SPIE.5904...21B,2010SPIE.7739E.123G}) applications.  To date, our group has produced the gratings on large (up to 100 mm diameter and 30 mm thick) monolithic pieces of monocrystalline silicon and we shape the substrates by sawing, grinding, and polishing to form the final devices \cite{2010SPIE.7739E.146W}.  The next generation of astronomy and earth science spectrographs, however, will need optical devices larger and more precise than what can be made with existing technology.  The proposed Giant Magellan Telescope \cite{2012SPIE.8444E..1HJ} Near Infrared Spectrograph \cite{2006SPIE.6269E.143J,2010SPIE.7735E..87L} (GMTNIRS) will need immersion gratings with thicknesses up to 60 mm and lengths up to 200 mm.  Substrates of the required size cannot fit into conventional processing equipment like reactive ion etchers and electron-beam lithography tools.  Large, thick Si substrates are also prohibitively expensive to experiment on.  Silicon direct bonding \cite{1986JAP....60.2987S,2012SPIE.8450E..2TV} offers advantages by allowing us to produce the diffraction grating on a thin (0.5$-$10 mm) low-mass Silicon wafer or puck that can later be bonded to a fine polished Si prism substrate.  Other groups producing Si diffractive optics are also experimenting with or using this approach (see \cite{2012SPIE.8450E..2TV} and \url{www.lumella.com}).

While it is possible that proper precautions can be made to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as ``tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Some authors \cite{1991JaJAP..30..615M, 1992JEMat..21..669M, 1989JaJAP..28L2141L, Mitani1990} have referred to large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  The gaps have also been referred to as voids\cite{2000RScI...71.1869G}, interfacial voids, or gaps.  In this paper we will reserve the term ``IR bubbles'' for gaps that are large enough to be detected directly in conventional IR imaging.  The term ``gap'' will apply more broadly to any finite size gap.

Finite gaps will degrade the performance of a final device, either through optical aberrations or transmission loss.  The effect of the gap size on the IR transmission is the topic of this paper.  The effect of the gap on optical aberrations is beyond the scope of this paper.

A gap might also degrade the mechanical integrity of a final compound optic.  In practice, the part must survive thermal shocks and shake tests.  The gap size scales we can probe with our optical technique are comparable to the range over which molecular (van der Waals) forces are acting.  It is conceivable that the strength of the bond energy can be constrained from the \emph{in situ} measurement of the separation and that therefore our optical technique can provide not only a measure of the transmission but also an indication of the mechanical robustness of the bond.  By measuring both the spatial and axial extent of any gaps, the new metrology technique informs revisions of the bonding strategy that can improve both the optical and mechanical performance.  We do not investigate the mechanical robustness further.

There are two main solutions for reducing risk attributable to gaps.  The first approach is to prevent gaps from forming in the first place, specifically by removing all possible contaminants from the environment before bonding.  Typically this approach involves disciplined handling protocols, and rigorous multi-stage cleaning steps.  The reason the approach is so hard is that hydrocarbons can serve as catalysts for gap formation\cite{1991JaJAP..30..615M}, and that these hydrocarbons are common in Si wafer handling equipment.  Unconventional micro-cleanroom approaches have been effective at meeting the goal of bubble-free bonding\cite{1989JaJAP..28L2141L}.  The second approach is to accept the likelihood that gaps will occur, and take steps to remove them.  The most cited way to get rid of gaps is to anneal them\cite{Mitani1990}.  We describe the current understanding of both approaches in Section \ref{secHistory}.

Whatever technique one uses to remove or reduce the effects of gaps, it is essential to have a quantitative method by which to assess accurately the optical impact of the gaps before and after treatment.  This paper introduces a technique for directly measuring gaps at silicon-silicon interfaces.  Mitani and G\"oselle review other techniques for measuring and mapping gaps\cite{1992JEMat..21..669M}.  One key technique is identifying IR Newton rings from IR imaging\cite{Mitani1990}.  The number and morphology of Newton rings yields information about the lateral and axial extent of large gaps. Specifically, dark-light fringes manifest when the gap is comparable to the vacuum wavelength or larger.  Our technique is IR interference in the limit of small gaps down to 20 nm or less.

There are other techniques for mapping gaps.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers.  In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

Here we describe an application of the IR interference technique with new precision instrumentation in the near infrared to measure the axial extent of gaps at the few nanometer size scale.  In this paper we show that gaps as small as 35 nm cause an unacceptable degradation in transmission performance.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=.8\columnwidth]{figs/20130911_VS21posM1}}
\caption{IR transmission spectrum of VS21 (M1) showing consistent gap\label{VS21specm1}.  Example spectrum through part VS21 at a position through the mesh gap of known dimensions.}
\end{figure}

>>>>

\subsection{Why Silicon is good}
Crystalline silicon is an excellent optical material for the infrared. Si has a high refractive index ($\sim$3.4, Frey et al. 2006) and superb transmission from slightly longer than the equivalent wavelength of its band gap ($\sim$1.15 $\mu$m) to about 6.5 $\mu$m (REF) and in the far-IR and submillimeter (REF).  It also has adequate transmission  (cite loss per cm$^{-1}$) for many purposes in the 18-30 $\mu$m region as well (REF).  There is an extensive industrial infrastructure for the production of ultrapure Si and for polishing and lithographic patterning and etching of optical parts.

\subsection{Why bonding is good and AR Coating is bad}
As with a number of glasses and crystalline optical materials, it is possible to form strong physical bonds between two planar or conformal silicon surfaces.  One advantage of this direct Si-Si bonding of macroscopic optical parts is that it facilitates the manufacture of complex optical subsystems, for example double-convex aspheres, combinations of lenses, prisms, and transmission gratings, and pairs of gratings oriented at right angles \emph{cite the Dutch group, one of our papers with cross-dispersed but not contacted grisms, and Lumella here}.  A second important advantage is that a sufficiently close bond is optically lossless, an important consideration not only for its effect on throughput but also because it can eliminate concerns about optical ghosts.  The remaining advantages of bonding stem from various difficulties encountered in antireflection coating a high index material in the near and mid-infrared:  There is a limited choice of index-matching materials, in particular at longer wavelengths.  Systems often are used at cryogenic temperatures and considerations of differential thermal expansion, brittleness and hygroscopic properties can restrict the choices even further. It is hard to get good antireflection performance across broad spectral ranges and at a range of incidence angles.

\subsection{Silicon Bonds- history and techniques}

There are several good reviews of Si bonding that include accounts of the history of the field and a description of process alternatives, metrology, and problems (Goesele and Tong, Mastreika et al.  2014).  The foundation paper on Si bonding came in 1986[14]. After that, there was a sequence of papers that used IR imaging to look for evidence of defects in the bond. Stengl et al. (1988) and Goesele et al. (1995) [23, 24] directly monitored the bond propagation with IR imaging. Lehman et al. 1989 and Goesele et al. [19, 23] demonstrated gap-free bonding in a micro- cleanroom, in which the bonding process occurs immediately after a rinse-spin-vacuum cycle inside of a vacuum enclosure.  Other authors have investigated the growth of bubbles at bond with time and the elimination of some or all of these bubbles after annealing bonded parts at temperatures of 800-1100 C (REFS including looking at some from the Mastreika paper).

The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy[25], which lent credibility to the idea of distinct temperature phases and evidence for Si?H bonds[26]. A robust picture of the chemistry as a function of anneal temperature was ultimately fleshed out[27, 28]. Other variables potentially affecting bond strength were studied, for example surface roughness[29], surface topography [30], surface preparation[31], annealing time [32], ambient pressure and substrate thickness [33, 34], and moisture and defects [35]. Notably, [29] found that surface roughness above 1.3 nm RMS results in poor bonding quality in their sample of wafers etched with Argon. Results from studies of surface topography are inconclusive[30].

\subsection{Requirement:  Do better than coatings averaged over the surface}

If Si-Si bonds are to be a useful part of the optical designer's toolbox, the bonds must have transmission losses similar to or lower than those of two antireflection coated surfaces over the operating wavelength (typically $< 2$\%).  If the optical layout does not eliminate pupil ghosts from the flat Si surfaces, there are additional requirements on the reflectivity of individual bubbles in the bond. While completely bonded Si parts will be completely lossless at the interface, problems on different scales with different physical origins can lead to less-than-perfect bonds.  On the smallest scales, imperfect cleaning can leave small voids where particulates hold the bonding planes apart or where hydrocarbons can serve as catalysts for gap formation[17].  Even in the best of cleanrooms, particles at the optically significant 10-20 nm scale can be present in relatively high abundance. Figure 5 of Cooper 1986[16] shows the surface flux (particles cm$^2$ s$^{-1}$) vs. particle size ($\mu$m). There will be roughly 104 times more 10 nm sized particles than 1 ?m sized particles.  On larger scale, gaseous hydrogen, which is a byproduct of the  hydrophilic bonding process, can migrate into the bond and form local bubbles (REFS).  Most of the bonding literature concerns itself with wafer bonding.  In bonding wafers to each other or to thicker optically polished Si disks, one or both of the partners can conform as the bonding front propagates.  When bonding two thicker and more rigid substrates, differences in the shapes of the polished surfaces can potentially lead to topological inconsistencies that prohibit some areas from bonding.

\subsection{Existing detection schemes}
Verification that an optical bond meets its requirements and possible diagnosis of problems in the bonding technique require an accurate assessment of the geometry and lateral and axial dimensions of any defects.  Since silicon is opaque in the optical, the most straightforward technique for detecting defects in the bond is to image it in the infrared.  Fresnel losses will cause unbonded regions to appear darker than regions where the bond is perfect and the transmission is 100\%.  In cases where the axial extent of the gap is a wavelength or more, Newton rings appear in the image of the gap [20].  X-ray topography [18, 22] exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but not the axial extent itself. In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but produces limited information about the axial dimension [21].

\subsection{Technique}
We describe here a new technique for measuring the axial extent of gaps, founded upon the ability to measure transmission as a function of wavelength with high accuracy.   Stock near-infrared spectrophotometer like the Cary5000 from Agilent offer 0.2\% precision and good repeatability.  We use the wavelength dependence of the transmission and a model of the gap as a closely-spaced Fabry-Perot to measure axial gaps down to a few nanometers in size.  In order to verify the applicability and accuracy of the technique, we have used microlithography to create small artificial gaps of known size in wafers that we then bond and measure.  Nowhere in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that was not widely available until this decade.

\subsection{What level do you need to measure to}
For all types of gaps and bubbles, our technique needs to provide an accurate assessment of the transmission loss at the operating wavelength.  For finite-sized bubbles, measurement of the axial dimension helps us to assess the effects of high-temperature annealing on removal of hydrogen bubbles.  A wavelength-dependent transmission variation can also indicate the presence of large numbers of gaps that are too small to resolve spatially, as one might expect if there are many very small contaminants present along the interface.  To meet these diagnostic goals, our technique must be able to characterize gaps with losses of a few percent at the operating wavelength.  A 2\% loss at 1.15 $\mu$m corresponds to a gap of XX nm at normal incidence and the gap size for a loss of this magnitude scales linearly with wavelength.  We therefore need to be able to measure transmission as a function of wavelength to a fraction of a percent, including all systematic errors, for devices operating near 1 $\mu$m and to a somewhat looser standard for devices operating at longer wavelengths.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=.8\columnwidth]{figs/20130911_VS21posM1.pdf}}
\caption{IR transmission spectrum of VS21 (M1) showing consistent gap\label{VS21specm1}.  Example spectrum through part VS21 at a position through the mesh gap of known dimensions.}
\end{figure}
